Reviewer Comments for Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0

### 2019-05-07 2019_04_22_Reviewer_Comments.Rmd
adding sup figures, editing tables, small typo fixes

1. Figure 4 is not readable. If this is going to be useful illustration of potential, it would be better to do each of the 5 subparts as a separate figure. Similarly, Figure 5 would be better separated into 2 figures with increased readability.
2. Minor typos in the manuscript, but otherwise clearly written.
```
Move Figure 4 to supplemental figure?

Split Figure 5 into two figures?

@@

All quantification of expression happens at the transcript level. As most researchers think in terms of gene expression, we collapse all of the transcript level information information into a gene. Very briefly, the tximport tool (https://f1000research.com/articles/4-1521/v2) is used to collapse expression from each transcript of a gene is averaged, with scaling for the length of the transcript.

We have

@@

Number check!

```
3) Line 179: "To create a high quality final dataset across the 1311 initial samples (suppl fig and table 2)". Are these "1311 initial samples" the same as "371 human eye tissue samples and 848 GTEx" described on line 158-159. If so, 371+848 ≠ 1311? 
```
Number check!
```
4) In Figure 3, the colors of ESC and lens, retina and RPE are too close to each other, and hard to be distinguished. More distinctive, different colors would help. What is the total number of clusters? A table listing all the clusters and the composition of each cluster should be included as a supplemental table. 
```
Number check!
```
5) Line 208, ".... Then genes (or transcripts), then tissues." The "genes" should be "gene IDs by official gene name". The tables derived from "expression" search (e.g. Fig.4A) should be further annotated, e.g. define what are "decile", "log2(PMT+1)", etc. 
```
HGNC gene names
```
6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.
```

@@

While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some empirical evidence that these types of tools may overinflate confidence in group differences (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/).

@@

```
4. Further information about the quality control and filtering applied to the RNA-seq data would be helpful to include in the manuscript. For example, how were the parameters for filtered by a mapping rate of 40% per sample and median read count >2 across all genes per sample selected? Also, basic statistics related to read mapping such as min, max, average, 25 quartile, 75 quartile of mapping should be included. 
```

Parameters selected by checking distributions and picking the parameter that removes the left tail.

Good idea - add those statistics

@@

In principle we agree. We were not expecting use the t-SNE parameters so heavily. We were surprised in our first paper that the t-SNE worked so well in clustering samples by tissue (though this may be overfitting ** ADD LIMITATION to manuscript**).

We are using the euclidean distance in a *qualitative* manner. This "cut-off" distance we select to remove samples is intended to only remove individual samples that are very far away like tissues. Nearly all of these outliers are not forming an independent group, but rather are near **other** tissues and thus are likely to be tube mislabels / sample extraction errors. Another possibility are samples which had technical issues with sequencing that still passed the basic QC metrics cutoffs.

That said ** TRY THE COOK'S DISTANCE SUGGESTION **
```

@@

** DO POWER CALCULATION **

library(ssizeRNA)
library(edgeR)
library(tidyverse)
tpm <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_gene_TPM_03.tsv.gz')
metadata <- read_tsv('https://hpc.nih.gov/~mcgaugheyd/eyeIntegration/2019_metadata_03.tsv.gz')

group <- metadata %>% 
  filter(sample_accession %in% names(tpm)) %>% 
  select(sample_accession, Tissue) %>% 
  unique() %>% 
  pull(Tissue)
d <- DGEList(tpm[,2:ncol(tpm)], group = groups)
d <- calcNormFactors(d)
d <- estimateCommonDisp(d)
power <- c()
for (i in seq(5,100,5)){
  power <- c(power, check.power(nGenes = 37000, pi0 = 0.6, i, d$counts[,1], d$tagwise.dispersion, 2, up = 0.5,  replace = TRUE, fdr = 0.05, sims = 2)$pow_qvalue_ave)}

data %>% ggplot(aes(x=V1, y=power)) + geom_line() + theme_minimal() + labs(x='Number of Samples in Each Group', y = 'Power to Detect 1 log2(FC)') + scale_x_continuous(breaks = seq(10,100,10), labels = seq(10,100,10))

The value of the full GTEx dataset comes in the power to identify expression quantitative trait loci (eQTL). For differential gene expression by tissue, far fewer samples are needed. A power calculation we ran gives us 80+% power to detect >= 1 log2(FC) gene expression differences with a 20 samples x 20 sample comparision. As we have many more eye samples, this is a underestimate of our power when doing eye vs (body) comparisons. We have added the power calculation as supplemental figure.

>>>>

1. Figure 4 is not readable. If this is going to be useful illustration of potential, it would be better to do each of the 5 subparts as a separate figure. Similarly, Figure 5 would be better separated into 2 figures with increased readability.
2. Minor typos in the manuscript, but otherwise clearly written.
```
We have reduce the number of panels in Figure 4 to two and moved the remaining panels to a supplementary figure.

Figure 5 has been split into two separate figures.

@@

Supplmental table 2 is unnecessarily confusing. There are no differences between the "gene type" whether at the gene or transcript level. We have added the Gencode description of the biotypes to the table and given the counts as "Gene Counts" or "Transcript Counts" for each biotype.

@@

We originally had 63 tissues - several were removed as they had an insufficient number of samples after quality control. I've updated the text for this figure legend to reflect the final number (58).

```
3) Line 179: "To create a high quality final dataset across the 1311 initial samples (suppl fig and table 2)". Are these "1311 initial samples" the same as "371 human eye tissue samples and 848 GTEx" described on line 158-159. If so, 371+848 ≠ 1311? 
```
1311 went into the pipeline. 30 GTEx and 62 eye samples were removed for a variety of QC reasons (see sup table). 371 + 848 + 30 + 62 = 1311. This was made more clear in the text. ** PUT IN NEW QUOTE AFTER KNITTING WORD DOCUMENTS** 
```
4) In Figure 3, the colors of ESC and lens, retina and RPE are too close to each other, and hard to be distinguished. More distinctive, different colors would help. What is the total number of clusters? A table listing all the clusters and the composition of each cluster should be included as a supplemental table. 
```
Updated colors.

There are 33 total clustesr. We have added a supplementary table summarizing the Sub Tissue composition of each one.

tsne_50 %>% group_by(Cluster, Sub_Tissue) %>% summarise(Count = n()) %>%  mutate(Sub_Tissue = trimws(Sub_Tissue))  %>% rename(`Sub Tissue` = Sub_Tissue)  %>% flextable::regulartable()
```
5) Line 208, ".... Then genes (or transcripts), then tissues." The "genes" should be "gene IDs by official gene name". The tables derived from "expression" search (e.g. Fig.4A) should be further annotated, e.g. define what are "decile", "log2(PMT+1)", etc. 
```
HGNC gene names
```
6) Since this dataset and web app is built in the way that "future sample updates can be streamlined in with less effort", perhaps advice on how future deep sequencing and single cell seq data should be organized for streamline updates should be added to the discussion.
```

@@

While  ComBat is an extremely popular and powerful tool to remove batch effects, I am hesitant to use as there is some evidence that these types of tools may overinflate confidence in group differences (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4679072/).

@@

```
4. Further information about the quality control and filtering applied to the RNA-seq data would be helpful to include in the manuscript. For example, how were the parameters for filtered by a mapping rate of 40% per sample and median read count >2 across all genes per sample selected? Also, basic statistics related to read mapping such as min, max, average, 25 quartile, 75 quartile of mapping should be included. 
```
Parameters selected by checking distributions and picking the parameter that removes the left tail. We have added mapping rate as a supplementary figure and edited the text in the methods accordingly.

@@

I wrote a long explanation of how what we are doing is reasonable, but then thought some more and now have come around to your position. While this t-SNE / euclidean distance cutoff approach works on a empirical level, it is a poor idea.

We have implemented the outlier-by-correlation-to-group approach you suggested (Nature Genetics 46: 430, 2014).
```

@@

The value of the full GTEx dataset comes in the power to identify expression quantitative trait loci (eQTL). For differential gene expression by tissue, far fewer samples are needed. A power calculation we ran gives us 80+% power to detect >= 1 log2(FC) gene expression differences with a 20 sample x 20 sample comparision. We would reach a power of 90% with 30 samples, so we have added 10 more samples to each GTEx tissue group. As we have many more eye samples, this is likely an underestimate of our power when doing eye vs (body) comparisons.

We have added the power calculation as supplemental figure.

While it would be nice in *principal* to just use the entire GTEx dataset, there are very practical limiations on why this is inconvenient - the size of the full GTEx datasets, raw, is around 1 petabyte of data, which is very slow to move around and do any compute on. Also this amount of storage space brushes against the limits of our compute cluster.

As for the Ratnapriya/Swaroop data, we mention it briefly in our introduction, but we hadn't added the data to eyeIntegration as it became available shortly after our last data pull.

We have surveyed the SRA / GEO again and have added new studies (including the Ratnapriya/Swaroop) project data to eyeIntegration.

As the AMD data they collected are likely to be of interest, we have made an exception to our "normal tissue" policy and have added the different AMD stages as selectable tissue options.
